---
title: "Where are the opioids going?"
subtitle: "An exploratory data analysis on Medicare Part D claims information"  
author: 
  - "Q Quaye"
  - "Shiron"
  - "Tolu"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
      
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1381B0",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)

#loading packages
library(tidyverse)
library(ggrepel)

#loading medicare data

medicare <- read_csv("http://www.stat.cmu.edu/cmsac/sure/2022/materials/data/health/eda_projects/medicare_partd_presc_claims.csv")

#loading state populations data

populations_2019 <- read_csv("/Users/Akwelle/Desktop/SURE 22/SURE EDA Project/state-populations-2019.csv")

populations_2019 <- as.vector(populations_2019$population)

#normalizing claims by population

normalizedClaims <- medicare %>%
  group_by(state = State, opioid = OpioidFlag) %>%
  summarize(sumNumberClaims = sum(NumberClaims))

avgNumClaims$population <- populations_2019

avgNumClaims_n <- avgNumClaims %>%
  mutate(numClaimsPer10000 = sumNumberClaims / population * 10000)

#filtering for only opioid prescription claims

opioidOnly <-  avgNumClaims_n %>%
  filter(opioid == "Opioid")

```
---
## Medicare Part D Prescription Claims Data

- 51,695 data points proportionally sampled across 50 US States + DC about drugs prescribed by health care providers

--

.pull-left[
  ### Categorical Variables:
  - Provider's name, age, city, state, and specialty
  - Drug name, whether or not the drug is an opioid
  - 
]

--

.pull-right[
  ### Continuous Variables:
  - drug cost
  - number of drug fills, number of days' supply
  - number of total drug claims, number of drug claims for those aged 65+
  
]
---

class: left

## Hypothesis: States with smaller populations prescribe more opioids


---

class: full-scale-fig

# Are smaller states prescribing more opioids?

```{r, include=TRUE, fig.align='center', echo=FALSE}

opioidOnly %>%
  ggplot(aes(x=log(population), y=numClaimsPer10000, label=state)) +
  geom_text_repel(color="red", size=2) +
  scale_color_viridis_c() +
  labs(x = "log(State Population)", y="Opioid Prescription Claims per 10,000 residents",
       main="Opioid Prescription Claims by State") +
  geom_point() +
  theme_minimal() +
  theme(axis.title = element_text(size=8))


```

--

### No.

---

- States with the lowest population have some of the lowest claims for opioid medications in the United States

--

- The state with the highest number of opioid prescriptions (WV) does have a smaller population

--

- States with higher populations (CA, TX) have lower levels of opioid prescriptions



